### 2024-06-18 basic-statistics.qmd
20240618 - misconceptions about p-values

$$
\bar{X} = \frac{\sum X_i}{n} = \frac{x_1 + x_2 + ... + x_n}{n}
$$ {#eq-IQR}

@@

To estimate a combination of indices of center, spread, and shape of the data, you can use the following code:

@@

@@

In science, we typically consider an effect to be *statistically significant* if the *p*-value is less than .05.
That is, there is a small chance (5%; or 1 in 20 times) that we would observe an effect at least as extreme as the effect observed, if the null hypothesis were true.
If the *p*-value is less than .05, we reject the null hypothesis ($H_0$) that there was no difference or association.
Thus, we conclude that there was a significant (non-zero) difference or association.
If the *p*-value is greater than .05, we fail to reject the null hypothesis; the difference/association was not significant.
Thus, we do not have confidence that there was a difference or association.
However, we do not accept the null hypothesis; it could be there we did not observe an effect because we did not have adequate power to detect the effect—e.g., if the effect size was small, the data were noisy, and the sample size was small and/or unrepresentative.

>>>>

$$
\bar{X} = \frac{\sum X_i}{n} = \frac{X_1 + X_2 + ... + X_n}{n}
$$ {#eq-IQR}

@@

To estimate multiple indices of center, spread, and shape of the data, you can use the following code:

@@

Nevertheless, to be conservative (i.e., to avoid false positive/Type I errors), many researchers use two-tailed *p*-values regardless whether their hypothesis is one- or two-tailed.

@@

In science, using what is called null-hypothesis significance testing (NHST), we typically consider an effect to be *statistically significant* if the *p*-value is less than .05.
That is, there is a small chance (5%; or 1 in 20 times) that we would observe an effect at least as extreme as the effect observed, if the null hypothesis were true.
So, you might expect around 5% of tests where the null hypothesis is true to be significant just by chance.

If the *p*-value is less than .05, we reject the null hypothesis ($H_0$) that there was no difference or association.
Thus, we conclude that there was a significant (non-zero) difference or association.
If the *p*-value is greater than .05, we fail to reject the null hypothesis; the difference/association was not significant.
Thus, we do not have confidence that there was a difference or association.
However, we do not accept the null hypothesis; it could be there we did not observe an effect because we did not have adequate power to detect the effect—e.g., if the effect size was small, the data were noisy, and the sample size was small and/or unrepresentative.

Interactive visualizations by Kristoffer Magnusson on *p*-values and null-hypothesis significance testing are below:

- <https://rpsychologist.com/pvalue/> (archived at <https://perma.cc/JP9F-9ZVY>)
- <https://rpsychologist.com/d3/pdist/> (archived at <https://perma.cc/BE96-8LSJ>)
- <https://rpsychologist.com/d3/nhst/> (archived at <https://perma.cc/ZU9A-37F3>)

Twelve misconceptions about *p*-values [@Goodman2008] are in @tbl-pValueMisconceptions.

| Number                                   | Misconception
|:----------------------------------------------------|:------------|
| 1 If $p = .05$, the null hypothesis has only a 5% chance of being true.
| 2 A nonsignificant difference (eg, $p > .05$) means there is no difference between groups.
| 3 A statistically significant finding is clinically important.
| 4 Studies with $p$-values on opposite sides of .05 are conflicting.
| 5 Studies with the same $p$-value provide the same evidence against the null hypothesis.
| 6 $p = .05$ means that we have observed data that would occur only 5% of the time under the null hypothesis.
| 7 $p = .05$ and $p < .05$ mean the same thing.
| 8 $p$-values are properly written as inequalities (e.g., "$p \le .05$" when $p = .015$)
| 9 $p = .05$ means that if you reject the null hypothesis, the probability of a Type I error is only 5%.
| 10 With a $p = .05$ threshold for significance, the chance of a Type I error will be 5%.
| 11 You should use a one-sided $p$-value when you don't care about a result in one direction, or a difference in that direction is impossible.
| 12 A scientific conclusion or treatment policy should be based on whether or not the $p$-value is significant.

Twelve Misconceptions About *p*-Values from @Goodman2008. Goodman also provides a discussion about why each is false. {#tbl-pValueMisconceptions}
